Constructing Biological Regulatory Networks from Process Hitting models

### 2013-08-03 parts-intro.tex
all source files in the same folder (TCS submission)

In order to address the formal analysis of dynamical properties within very large BRNs, we recently
introduced in \cite{PMR10-TCSB} a new formalism, named the \emph{Process Hitting} (PH), to model
concurrent systems having components with a few qualitative levels.
A PH model describes, in an atomic manner, the possible evolutions of a process (representing one
component at one level) triggered by the hit of at most one other process in the system.
This framework can be seen as a special class of formalisms like Petri Nets or Communicating Finite
State Machines, where the arity and the effect of synchronizations between components is
restricted.
Thanks to the particular structure of interactions within a PH, very efficient static analysis
methods have been developed to over- and under-approximate reachability properties making tractable
the formal analysis of the qualitative dynamics of BRNs with hundreds or thousands of components,
which was impossible with other state-of-the-art approaches \cite{PMR12-MSCS,PAK13-CAV,FPMR13-CS2Bio}.

@@

Firstly, we derive rules that, given a PH model, infer the IG corresponding to the dynamics of the
encoded BRN.
The obtained IG relates only components that have actual influence in the dynamics.
This typically results into IGs that contain less interaction than an hypothetical starting prior
IG, as the specification of component evolutions may reveal non-functional regulations.
Based on the derived IG, various static analysis allows to conclude on global properties
of the system dynamics (see \cite{PR11-SASB} for a short survey).
The most known are the Thomas' conjectures (that have been proved since,
e.g.~\cite{RRT08,RiCo07,Richard2010378}
for Boolean and discrete networks),
which relate the absence of
positive cycle to the impossibility for multi-stationarity (distinct attractors),
and the absence of negative cycle to the impossibility for oscillatory behaviours.

Second, we tackle the characterisation of Thomas' parameters that are compatible with a given PH
model, \ie that fully specify the evolution of components while respecting the cooperations
specified in PH.
More formally, the resulting BRN dynamics is ensured to be included in the PH dynamics: any
transition in the (asynchronous) Thomas model exists in the PH model.
In practice, the number of compatible Thomas parameters can be huge for large networks where only a
few cooperations are specified.
This highlights the fact that the PH is very suited to analyse dynamics of large-scale networks
where the knowledge on cooperations is often incomplete:
instead of dealing with a possibly non-tractable number of Thomas models, one can already capture
precise dynamical features using a single PH models which gathers all the plausible dynamics.

@@

Our work is related to the approach of \cite{Khalis09} which relies on temporal logic, and \cite{20646302,DBLP:conf/ipcat/CorblinFTCT12} which also uses constraint programming.
Both aim at determining a class of models which are consistent with available partial data on the regulatory structure and dynamical properties.
%These classes are built in order to infer properties common to some studied models.
%In our approach, we intend to focus on the Thomas' parameters inference
Our method is based on a model rather than on constraints, which allows to define some properties on the system structure (such as cooperations).
Furthermore, we claim that we are able to deal with larger biological networks.
Finally, it must be noticed that we are not interested in this paper in the derivation of one PH
from a BRN (which was previously described in \cite{PMR10-TCSB}) but, on the contrary, to finding
out a set of BRNs from one PH.

The contributions presented in this paper significantly extend and improve the preliminary results
presented in \cite{FPIMR12-CMSB}.
In addition of improving the efficiency and accuracy of the IG inference, we have added support for
unsigned regulations, that are regulations being positive or negative depending on a particular
context.
This allows to apply our methodology to a wider class of models.
Furthermore, some parts have been completely rewritten in order to explain more precisely the efficiency of the ASP implementation,
and give a new detailed biological application of our work.

>>>>

In order to address the formal analysis of dynamical properties within very large BRNs, we recently
introduced in \cite{PMR10-TCSB} a new formalism, named the \emph{Process Hitting} (PH), to model
concurrent systems having components with a few qualitative levels.
A PH model describes, in an atomic manner, the possible evolutions of a process (representing one
component at one level) triggered by the hit of at most one other process in the system.
This framework can be seen as a special class of formalisms like Petri Nets or Communicating Finite
State Machines, where the arity and the effect of synchronizations between components are
restricted.
Thanks to the particular structure of interactions within a PH, very efficient static analysis
methods have been developed to over- and under-approximate reachability properties making tractable
the formal analysis of the qualitative dynamics of BRNs with hundreds or thousands of components,
which was impossible with other state-of-the-art approaches \cite{PMR12-MSCS,PAK13-CAV,FPMR13-CS2Bio}.

@@

Firstly, we derive rules that, given a PH model, infer the IG corresponding to the dynamics of the
encoded BRN.
The obtained IG relates only components that have actual influence in the dynamics.
This typically results into IGs that contain less interactions than an hypothetical starting prior
IG, as the specification of component evolutions may reveal non-functional regulations.
Based on the derived IG, various static analysis allows to conclude on global properties
of the system dynamics (see \cite{PR11-SASB} for a short survey).
The most known are the Thomas' conjectures (that have been proved since,
e.g.~\cite{RRT08,RiCo07,Richard2010378}
for Boolean and discrete networks),
which relate the absence of
positive cycle to the impossibility for multi-stationarity (distinct attractors),
and the absence of negative cycle to the impossibility for oscillatory behaviors.

Second, we tackle the inference of Thomas' parameters that are compatible with a given PH
model, \ie that fully specify the evolution of components while respecting the cooperations
specified in PH.
More formally, the resulting BRN dynamics is ensured to be included in the PH dynamics: any
transition in the (asynchronous) Thomas model exists in the PH model.
In practice, the number of compatible Thomas parameters can be huge for large networks where only a
few cooperations are specified.
This highlights the fact that the PH is very suitable to analyze dynamics of large-scale networks
where the knowledge on cooperations is often incomplete:
instead of dealing with a possibly non-tractable number of Thomas models, one can already capture
precise dynamical features using a single PH model which gathers all the plausible dynamics.

@@

Our work is related to the approach of \cite{Khalis09} which relies on temporal logic, and \cite{20646302,DBLP:conf/ipcat/CorblinFTCT12} which also uses constraint programming.
Both aim at determining a class of models which are consistent with available partial data on the regulatory structure and dynamical properties.
%These classes are built in order to infer properties common to some studied models.
%In our approach, we intend to focus on the Thomas' parameters inference
Our method is based on a model rather than on constraints, which allows to define some properties on the system structure (such as cooperations).
Furthermore, we claim that we are able to deal with larger biological networks.
Finally, it must be noticed that we are not interested in this paper in the derivation of one PH
from a BRN (which was previously described in \cite{PMR10-TCSB}) but, on the contrary, in finding
out a set of BRNs from one PH.

The contributions presented in this paper significantly extend and improve the preliminary results
introduced in \cite{FPIMR12-CMSB}.
In addition to the improvement of the efficiency and accuracy of the IG inference, we have added support for
unsigned regulations, that are regulations being positive or negative depending on a particular
context.
This allows to apply our methodology to a wider class of models.
Furthermore, some parts have been completely rewritten in order to explain more precisely the efficiency of the ASP implementation,
and give a new detailed biological application of our work.

### 2013-08-03 parts-ph.tex
all source files in the same folder (TCS submission)

While the construction of cooperation in PH allows to encode any boolean functions
between cooperating processes \cite{PMR10-TCSB}, it is worth noticing they introduce a temporal
shift in their application.
This allows the existence of interleaving of actions leading to a cooperative sort representing a
past sub-state of the presence of the cooperative processes.
The resulting behavior is then an over-approximation
of the realization of an instantaneous cooperation.

>>>>

While the construction of cooperation in PH allows to encode any boolean functions
between cooperating processes \cite{PMR10-TCSB}, it is worth noticing they introduce a temporal
shift in their application.
This allows the existence of interleaving of actions leading to a cooperative sort representing a
past sub-state of the presence of the cooperative processes.
The resulting behavior is then an over-approximation
of the realization of an instantaneous cooperation.
However, it is possible to obtain the exact behavior of the cooperations by using the semantics with prioritized actions defined in~\cite{FPMR13-CS2Bio}.
Although this new semantics is not further described here, the inferences given in \pref{sec:infer-IG} \& \ref{sec:infer-K} do ignore the temporal shift induced by the cooperative sorts.
